---
title: "Advanced R Programming Assignment"
output: html_document
---

##Introduction

The overall purpose of this assessment is to evaluate your ability to apply functional programming and object oriented programming concepts in R. There are two parts to this assignment. The first part compares different functional programming techniques and benchmarks their performance. The second part uses object-oriented programming techniques to define a class to represent longitudinal data and provide a set of functions for interacting with such data.

###Review criteria

For both parts of this assignment you are required to write R code and to produce some analysis or output. You will be assessed on both the R code and the quality/accuracy of the output.

Please note that you should never execute someone else's R code. Evaluation of R code should be done via inspecting the code itself and by making your best assessment as to whether the code satisfies the rubric provided. 

For each part you will need to prepare a script file and an output file. The script file contains your code (i.e. functions, definitions) and the output file contains the output/analysis obtained by running your code. You will be evaluated on both the quality of the code and the appropriateness of the output.

###Step-By-Step Assignment Instructions

####Part 1: Factorial Function

The objective of Part 1 is to write a function that computes the factorial of an integer greater than or equal to 0. Recall that the factorial of a number n is n * (n-1) * (n - 2) * … * 1. The factorial of 0 is defined to be 1. Before taking on this part of the assignment, you may want to review the section on Functional Programming in this course.

For this Part you will need to write four different versions of the Factorial function:

1. Factorial_loop: a version that computes the factorial of an integer using looping (such as a for loop)
2. Factorial_reduce: a version that computes the factorial using the reduce() function in the purrr package. Alternatively, you can use the Reduce() function in the base package.
3. Factorial_func: a version that uses recursion to compute the factorial.
4. Factorial_mem: a version that uses memoization to compute the factorial.

After writing your four versions of the Factorial function, use the microbenchmark package to time the operation of these functions and provide a summary of their performance. In addition to timing your functions for specific inputs, make sure to show a range of inputs in order to demonstrate the timing of each function for larger inputs.

In order to submit this assignment, please prepare two files:

1. factorial_code.R: an R script file that contains the code implementing your classes, methods, and generics for the longitudinal dataset.
2. factorial_output.txt: a text file that contains the results of your comparison of the four different implementations.

####Part 2: Longitudinal Data Class and Methods

The purpose of this part is to create a new class for representing longitudinal data, which is data that is collected over time on a given subject/person. This data may be collected at multiple visits, in multiple locations. You will need to write a series of generics and methods for interacting with this kind of data.

The data for this part come from a small study on indoor air pollution on 10 subjects. Each subject was visited 3 times for data collection. Indoor air pollution was measured using a high-resolution monitor which records pollutant levels every 5 minutes and the monitor was placed in the home for about 1 week. In addition to measuring pollutant levels in the bedroom, a separate monitor was usually placed in another room in the house at roughly the same time.

Before doing this part you may want to review the section on object oriented programming.

The data are available as a CSV file and here:

[data.zip](https://d18ky98rnyall9.cloudfront.net/_257dbf6be13177cd110e3ef91b34ff67_data.zip?Expires=1485216000&Signature=Xvtys0sDyIwwmmAJaER6WhrdxuWaQG2uNJGjSkIOGaXzMyjaq3Mbg3l7cBtzsWJVDl~umUdjXYOvcAJy8yIu3eoZ2fD~VsF8VmIjM9bpAcDSPDFH3NNQJ2EX2LjtDcOBBhULo2j-K9FRsx3sq7VvC8GgZt3BoQHJdWbEdg6Q39k_&Key-Pair-Id=APKAJLTNE6QMUY6HBC5A)

The variables in the dataset are

- id: the subject identification number
- visit: the visit number which can be 0, 1, or 2
- room: the room in which the monitor was placed
- value: the level of pollution in micrograms per cubic meter
- timepoint: the time point of the monitor value for a given visit/room

You will need to design a class called “LongitudinalData” that characterizes the structure of this longitudinal dataset. You will also need to design classes to represent the concept of a “subject”, a “visit”, and a “room”.

In addition you will need to implement the following functions

- make_LD: a function that converts a data frame into a “LongitudinalData” object
- subject: a generic function for extracting subject-specific information
- visit: a generic function for extracting visit-specific information
- room: a generic function for extracting room-specific information

For each generic/class combination you will need to implement a method, although not all combinations are necessary (see below). You will also need to write print and summary methods for some classes (again, see below).

To complete this Part, you can use either the S3 system, the S4 system, or the reference class system to implement the necessary functions. It is probably not wise to mix any of the systems together, but you should be able to compete the assignment using any of the three systems. The amount of work required should be the same when using any of the systems.

For this assessment, you will need to implement the necessary functions to be able to execute the code in the following script file:

*oop_output.R*

```{r eval=FALSE}
## Read in the data
library(readr)
library(magrittr)
source("oop_code.R")
## Load any other packages that you may need to execute your code

data <- read_csv("data/MIE.csv")
x <- make_LD(data)
print(class(x))
print(x)

## Subject 10 doesn't exist
out <- subject(x, 10)
print(out)

out <- subject(x, 14)
print(out)

out <- subject(x, 54) %>% summary
print(out)

out <- subject(x, 14) %>% summary
print(out)

out <- subject(x, 44) %>% visit(0) %>% room("bedroom")
print(out)

## Show a summary of the pollutant values
out <- subject(x, 44) %>% visit(0) %>% room("bedroom") %>% summary
print(out)

out <- subject(x, 44) %>% visit(1) %>% room("living room") %>% summary
print(out)
```

The output should appear similar to the output in the following file:

*oop_output.txt*

```{r eval=FALSE}
## Read in the data
library(readr)
library(magrittr)
source("oop_code.R")
## Load any other packages that you may need to execute your code

data <- read_csv("data/MIE.csv")
x <- make_LD(data)
print(class(x))
[1] "LongitudinalData"
print(x)
Longitudinal dataset with 10 subjects 
 
## Subject 10 doesn't exist
out <- subject(x, 10)
print(out)
NULL
 
out <- subject(x, 14)
print(out)
Subject ID: 14 
 
out <- subject(x, 54) %>% summary
print(out)
ID: 54 
  visit  bedroom       den living room    office
1     0       NA        NA    2.792601 13.255475
2     1       NA 13.450946          NA  4.533921
3     2 4.193721  3.779225          NA        NA
 
out <- subject(x, 14) %>% summary
print(out)
ID: 14 
  visit   bedroom family  room living room
1     0  4.786592           NA     2.75000
2     1  3.401442     8.426549          NA
3     2 18.583635           NA    22.55069
 
out <- subject(x, 44) %>% visit(0) %>% room("bedroom")
print(out)
ID: 44 
Visit: 0 
Room: bedroom 
 
## Show a summary of the pollutant values
out <- subject(x, 44) %>% visit(0) %>% room("bedroom") %>% summary
print(out)
ID: 44 
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
    8.0    30.0    51.0    88.8    80.0   911.0 
 
out <- subject(x, 44) %>% visit(1) %>% room("living room") %>% summary
print(out)
ID: 44 
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
   2.75   14.00   24.00   41.37   37.00 1607.00 
```

The output of your function does not need to match exactly, but it should convey the same information.
In order to submit this assignment, please prepare two files:

- oop_code.R: an R script file that contains the code implementing your classes, methods, and generics for the longitudinal dataset.
- oop_output.txt: a text file containing the output of running the above input code.

###Part 1: Factorial Function

Just some reminders:

- purrr:  Reduce - List or vector reduction iteratively combines the first element of a vector with the second element of a vector, then that combined result is combined with the third element of the vector, and so on until the end of the vector is reached. The function to be applied should take at least two arguments. Where mapping returns a vector or a list, reducing should return a single value.

- Recursion - Recursion is very powerful tool, both mentally and in software development, for solving problems. Recursive functions have two main parts: a few easy to solve problems called “basecases,”and then a case for more complicated problems where the function is called inside of itself.The central philosophy of recursive programming is that problems can be broken down into simpler parts, and then combining those simple answers results in the answer to a complex problem. 

- Memoization - Memoization stores values of each calculated value in a table so that once a number is calculated you can look it up  instead of needing to recalculate it.  Thuis can speed the performance of a function. 

```{r factorial}
Factorial_loop <- function(x=1) {
  stopifnot(x >= 0)
  if (x == 0) {
    return(1)
} else {
    output <- 1
    for (y in 1:x) {
      output <- output * y
    }
 return(output)
  }
}

#Note:  Below must use numeric.  as.integer creates an warning of integer overflow NAs
library(purrr)

Factorial_reduce <- function(x=1) {
  stopifnot(x >= 0)
  if (x == 0) {
    return(1)
} else {
    reduce(as.numeric(1:x), function(x,y) {
      x * y
    })
  }
}

#Recursion
Factorial_func <- function(x=1) {
  stopifnot(x >= 0)
  if(x == 0) {1}
else {x * Factorial_func(x-1)
  }
}

#Memoization
Factorial_mem <- function(x=1) {
     stopifnot(x >= 0)
     Fact_tbl<- c(1)
     Factorial_mem_in <- function(x=1) {
       if(!is.na(Fact_tbl[x])) {Fact_tbl[x]}
       
       else{
         Fact_tbl[x] <<- x * Factorial_mem_in(x-1)
         return(Fact_tbl[x])}
     }
}

library(microbenchmark)
##Comparison for n=10
microbenchmark(a <- Factorial_loop(10),
               b <- Factorial_reduce(10),
               c <- Factorial_func(10),
               d <- Factorial_mem(10))
##Comparison for n=100
microbenchmark(a <- Factorial_loop(100),
               b <- Factorial_reduce(100),
               c <- Factorial_func(100),
               d <- Factorial_mem(100))
```

###Part 2: Longitudinal Data Class and Methods

```{r classCreation}
setGeneric("print")
setGeneric("summary")
setGeneric("subject", function(x,...){
  standardGeneric("subject")
})
setGeneric("visit", function(x,...){
  standardGeneric("visit")
})
setGeneric("room", function(x,...){
  standardGeneric("room")
})

# LongitudinalData Class and Methods

setClass("LongitudinalData",
         representation(id = "numeric", 
                      visit = "numeric",
                      room = "character",
                      value = "numeric",
                      timepoint = "numeric"))

setMethod("print",
          c(x = "LongitudinalData"),
          function(x){
            paste("Longitudinal dataset with", length(unique(x@id)), "subjects")
          })

setMethod("subject",
          c(x = "LongitudinalData"),
          function(x,n){
            new("subject_class", id = x@id[x@id == n], visit = x@visit[x@id == n],
                room = x@room[x@id == n], value = x@value[x@id == n],
                timepoint = x@timepoint[x@id == n])
          })

# subject_class Class and Methods

setClass("subject_class",
         representation(id = "numeric", 
                      visit = "numeric",
                      room = "character",
                      value = "numeric",
                      timepoint = "numeric"))

setMethod("print",
          c(x = "subject_class"),
          function(x){
            if (length(unique(x@id)) > 0) {
              cat(paste("Subject ID:",unique(x@id)))
            } else {
              NULL
            }
          })

setMethod("summary",
          c(object = "subject_class"),
          function(object){
            new("subject_summary", id = object@id, visit = object@visit, 
                room = object@room, value = object@value)
          })

setMethod("visit",
          c(x = "subject_class"),
          function(x,n){
            new("visit_class", id = x@id[x@visit == n], visit = x@visit[x@visit == n],
                room = x@room[x@visit == n], value = x@value[x@visit == n],
                timepoint = x@timepoint[x@visit == n])
          })

# subject_summary Class and Methods

setClass("subject_summary",
         representation(id = "numeric", 
                        visit = "numeric",
                        room = "character",
                        value = "numeric"))

setMethod("print",
          c(x = "subject_summary"),
          function(x){
            cat(paste("ID:", unique(x@id)),"\n")
            as.data.frame(cbind(visit=x@visit,room=x@room,value=x@value),stringsAsFactors = FALSE) %>%
              mutate(value = as.numeric(value)) %>%
              group_by(visit,room) %>%
              summarise(avg = mean(value)) %>%
              spread(room,avg)
          })

# visit_class Class and Methods

setClass("visit_class",
         representation(id = "numeric", 
                        visit = "numeric",
                        room = "character",
                        value = "numeric",
                        timepoint = "numeric"))

setMethod("room",
          c(x = "visit_class"),
          function(x,n){
            new("room_class", id = x@id[x@room == n], visit = x@visit[x@room == n],
                room = x@room[x@room == n], value = x@value[x@room == n],
                timepoint = x@timepoint[x@room == n])
          })

# room_class Class and Methods

setClass("room_class",
         representation(id = "numeric", 
                        visit = "numeric",
                        room = "character",
                        value = "numeric",
                        timepoint = "numeric"))

setMethod("print",
          c(x = "room_class"),
          function(x){
            cat(paste("ID:", unique(x@id)),"\n")
            cat(paste("Visit:",unique(x@visit)),"\n")
            cat(paste("Room:",unique(x@room)))
          })

setMethod("summary",
          c(object = "room_class"),
          function(object){
            new("room_summary", id = object@id, value = object@value)
          })

# room_summary Class and Methods

setClass("room_summary",
         representation(id = "numeric", 
                        value = "numeric"))

setMethod("print",
          c(x = "room_summary"),
          function(x){
            cat(paste("ID:", unique(x@id)),"\n")
            summary(x@value)
          })

# other functions

make_LD <- function(x) {
  new("LongitudinalData", id = x$id, visit = x$visit,
      room = x$room, value = x$value, timepoint = x$timepoint)
}

```