---
title: "Mutate"
output: html_document
---

The mutate function in dplyr can be used to add new columns to a data frame or change existing columns in the data frame. As an example, I’ll use the worldcup dataset from the package faraway. To load this example data frame, you can run:
```{r message=FALSE, warning=FALSE}
library(dplyr)
library(faraway)
data(worldcup)
head(worldcup)
```
This dataset has observations by player, including the player’s team, position, amount of time played in this World Cup, and number of shots, passes, tackles, and saves. This dataset is currently not tidy, as it has one of the variables (players’ names) as rownames, rather than as a column of the data frame. You can use the mutate function to move the player names to its own column:
```{r}
worldcup <- worldcup %>% mutate(player_name = rownames(worldcup))
worldcup %>% slice(1:3)
```

You can also use mutate in coordination with group_by to create new columns that give summaries within certain windows of the data. For example, the following code will add a column with the average number of shots for a player’s position added as a new column. While this code is summarizing the original data to generate the values in this column, mutate will add these repeated summary values to the original dataset by group, rather than returning a dataframe with a single row for each of the grouping variables (try replacing mutate with summarize in this code to make sure you understand the difference).
```{r}
worldcup <- worldcup %>% group_by(Position) %>% mutate(ave_shots = mean(Shots)) %>% ungroup()
glimpse(worldcup)

#With Summarize
worldcup2 <- worldcup %>% group_by(Position) %>% summarize(ave_shots = mean(Shots)) %>% ungroup()
worldcup2 %>% head()
```

If there is a column that you want to rename, but not change, you can use the rename function. For example:

```{r}
worldcup %>% rename(Name = player_name) %>% slice(1:3)
```

